## Problem 3: Predicting sprint times

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
```
## Problem 3: Predicting sprint times

Loading the data into R
```{r}
sprinters <- read.csv("sprinters.csv")
```

a. The referenced paper only used data from the Olympics 2004 and before. Create a new dataset named `sprinters_orig` with only those observations.
    ```{r}
    sprinters_orig <- subset(sprinters, year < 2005 & olympics == 1)
    ```

b. Run the regressions
    ```{r}
    lm(time ~ year + women, data = sprinters_orig)
    lm(time ~ year * women, data = sprinters_orig)
    lm(time ~ year, data = filter(sprinters_orig, women == 1))
    lm(time ~ year, data = filter(sprinters_orig, women == 0))
    ```
    Interpret each regression. How are they similar or different in their 
    slopes? Plot each of these using the **texreg** package.
      In general, I believe sprinters' times will decrease for each increase in year, by 0.012 seconds, and the first regression suggests that women's times will be increasing with ech year, by 1.04 seconds. When considering the interaction between year and women, the directions remain the same for 'year' and 'women' but the slope changes dramatically for women - with each spint times increasing by 11 seconds with each year. However, the interaction year*women variable is negative. When the data is filtered to include only women, the slope of sprinters' times with each increase in year is more negative (-0.01584) than when the data is filtered to include only men (-0.01068) - suggesting that women's times will actually decrease over time at a faster rate than mens' times.


c. Plot the results of these regressions.
